\textbf{Seasonal Dynamics of Epiphytic Microbial Communities on Marine Macrophyte Surfaces}

### 2020-03-02 submission/manuscript.tex
Prist part of discussion.

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of eukaryotic,
chloroplast, mitochondrial and no relative sequences
(\autoref{supp-nseq_notus}). A total of 35 samples originating from
epiphytic archaeal and bacterial communities associated with surfaces of
the seagrass \emph{Cymodocea nodosa} and macroalga \emph{Caulerpa
cylindracea} were analyzed. In addition, 18 samples (one of the samples
was sequenced two times) originating from picoplankton archaeal and
bacterial communities in the surrounding seawater were also processed.
The number of reads per sample ranged between 8,407 and 77,465 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off that is a common
observation in high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences mentioned before reads were clustered into 28,726 different
OTUs at a similarity level of 97 \si{\percent}. Reads numbers were
normalized to the minimum number of sequences, 8,407
(\autoref{supp-nseq_notus}), through rarefaction resulting in 16,963
different OTUs that ranged from 365 to 2,019 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes of richness
and diversity the Observed Number of OTUs, Chao1, ACE, Exponential
Shannon (Jost, 2006) and Inverse Simpson were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{Caulerpa cylindracea} (invaded {[}Number of OTUs,
1,686.9 ± 135.8 OTUs{]} and noninvaded {[}Number of OTUs, 1,742.8 ±
152.4 OTUs{]}), middle vales for \emph{Cymodocea nodosa} (Number of
OTUs, 1,047.7 ± 212.1 OTUs) and lower values for picoplankton
communities in the surrounding seawater (Number of OTUs, 528.4 ± 144.5
OTUs) (\autoref{supp-calculators}). Seasonal changes did not show such
large dissimilarities. Seawater communities richness was stable during
the studied period with the exception of one sampling point in December
2017 when larger values were observed. \emph{Cymodocea nodosa}
communities showed a slow increase towards the end of the study, while
\emph{Caulerpa cylindracea} (invaded and noninvaded) communities were
characterized by slightly larger values in Spring and Summer in
comparison to Autumn and Winter (\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient were, respectively, calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from a
particular environment. The highest proportion of shared OTUs and
community was found between invaded and noninvaded \emph{Caulerpa
cylindracea} (Jaccard, 0.35; Bray-Curtis, 0.77), while lower shared
values were calculated between seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{Cymodocea nodosa}
and \emph{Caulerpa cylindracea} were approximately in the middle between
these two extremes. To assess seasonal changes in the proportion of
shared OTUs and communities the Jaccard's and Bray-Curtis Similarity
Coefficients were calculated between consecutive sampling points
(\autoref{shared}). Both coefficients showed similar trends. Temporal
proportional changes were more pronounced for seawater in comparison to
\emph{Cymodocea nodosa} and especially \emph{Caulerpa cylindracea}
associated communities (\autoref{shared}). To further disentangle the
environmental and seasonal community dissimilarity a Principal
Coordinates Analysis (PCoA) based on Bray-Curtis distances and OTU
abundances was applied. It showed a clear separation between planktonic
and surface associated communities (\autoref{pcoa}). In addition, a
separation of epiphytic bacterial and archaeal communities based on host
species was determined. This separation was further supported by ANOSIM
(R = 0.96, \emph{p} \textless{} 0.001). Seasonal changes of seawater
communities indicated a separation between Spring, Summer and
Autumn/Winter samples (ANOSIM, R = 0.64, \emph{p} \textless{} 0.001).
Epiphytic microbial communities associated with \emph{Cymodocea nodosa}
showed a similar pattern (ANOSIM, R = 0.54, \emph{p} \textless{} 0.01),
while communities from the surfaces of \emph{Caulerpa cylindracea}
indicated a non so strongly supported, as in previous cases, separation
between Summer and Autumn/Winter/Spring samples (ANOSIM, R = 0.30,
\emph{p} \textless{} 0.01) (\autoref{pcoa}).

The microbial taxonomic composition of both, macrophyte associated and
seawater communities, was dominated by \emph{Bacteria} (99.1 ± 2.1
\si{\percent}) over \emph{Archaea} (0.9 ± 2.1 \si{\percent})
(\autoref{community}). Three phyla were characteristic for
\emph{Archaea}: \emph{Crenarchaeota}, \emph{Nanoarchaeota} and
\emph{Thermoplasmatota}. Their contribution was larger only in seawater
communities during December 2017 (\autoref{community}). Higher relative
abundances of chloroplast related sequences were only observed in
surface associated communities, with higher values in Autumn/Winter
(37.2 ± 11.2 \si{\percent}) in comparison to Spring/Summer (20.9 ± 9.7
\si{\percent}) (\autoref{supp-chloroplast}). Generally, at higher
taxonomic ranks (phylum-class) epiphytic and seawater microbial
communities were composed of similar bacterial taxa. Seawater
communities were mainly comprised of \emph{Actinobacteriota},
\emph{Bacteroidota}, \emph{Cyanobacteria}, \emph{Alphaproteobacteria},
\emph{Gammaproteobacteria} and \emph{Verrucomicrobiota}. Communities
associated with \emph{Cymodocea nodosa} were consisted of same groups
with the addition of \emph{Planctomycetota} whose contribution was
higher in summer 2018. In addition, communities from invaded and
noninvaded \emph{Caulerpa cylindracea} were similar and characterized by
same groups as seawater and \emph{Cymodocea nodosa} communities with the
addition of \emph{Desulfobacterota} (\autoref{community}). Larger
differences between environments and host species could be observed at
lower taxonomic ranks (\autoref{cyano} -- \ref{desulfo}).

\emph{Cyanobacteria} related sequences were comprising, on average, 5.5
± 4.4 \si{\percent} of total sequences (\autoref{cyano}). Higher
proportions were found for \emph{Cymodocea nodosa} (16.4 ± 5.3
\si{\percent}) and \emph{Caulerpa cylindracea} (invaded {[} (7.7 ± 3.9
\si{\percent}){]} and noninvaded {[} (7.8 ± 2.4 \si{\percent}){]})
associated communities in autumn and for seawater communities in winter
(8.8 ± 7.4 \si{\percent}). Large taxonomic differences between surface
associated and seawater cyanobacterial communities were observed.
Seawater communities were mainly comprised of \emph{Cyanobium} and
\emph{Synechococcus}, while surface associated communities were
consisted of and sequences without known relatives within
\emph{Cyanobacteriales} and \emph{Cyanobacteriia} (\autoref{cyano}). In
addition, seasonal changes in surface associated communities were
observed with \emph{Cyanobacteriales} and \emph{Cyanobacteriia} no
relative sequences comprising larger proportions in autumn and winter
and \emph{Acrophormium}, \emph{Phormidesmis} and no relative
\emph{Nodosilineaceae} and in spring and summer (\autoref{cyano}).

Sequences classified as \emph{Bacteroidota} were comprising, on average,
19.2 ± 5.5 \si{\percent} of all sequences (\autoref{bactero}). Similarly
to \emph{Cyanobacteria}, large differences in the taxonomic composition
between seawater and surface associated communities were found
(\autoref{bactero}). The seawater community was characterized by the NS4
and NS5 marine groups, uncultured \emph{Cryomorphaceae}, uncultured
\emph{Flavobacteriaceae}, NS11-12 marine group, \emph{Balneola},
uncultured \emph{Balneolaceae} and \emph{Formosa}. In contrast, surface
associated communities were characterized by \emph{Lewinella},
\emph{Portibacter}, \emph{Rubidimonas}, no relative
\emph{Saprospiraceae}, uncultured \emph{Saprospiraceae}, no relative
\emph{Flavobacteriaceae} and uncultured \emph{Rhodothermaceae}. Surface
associated \emph{Bacteroidota} communities were very diverse as could be
observed in the the high proportion of taxa that grouped as other
\emph{Bacteroidota} (\autoref{bactero}).

On average, \emph{Alphaproteobacteria} were in comparison to other high
rank taxa the largest taxonomic group, comprising 29.2 ± 12.0
\si{\percent} of all sequences (\autoref{alpha}). In accordance to
previous taxa, high differences between seawater and surface associated
communities were observed. Picoplankton communities were composed mainly
of the SAR11 clade, AEGEAN-169 marine group, SAR116 clade, no relative
\emph{Rhodobacteraceae}, HIMB11 and OCS116 clade, while surface
associated communities were composed in high proportion of no relative
\emph{Rhodobacteraceae} and to a lesser degree of \emph{Pseudoahrensia},
no relative \emph{Alphaproteobacteria} and no relative
\emph{Hyphomonadaceae}. Representatives of no relative
\emph{Rhodobacteraceae} were comprising in on average 40.6 ± 23.2
\si{\percent} of all alphaproteobacterial sequences from the epiphytic
community (\autoref{alpha}).

Sequences related to \emph{Gammaproteobacteria} were comprising, on
average, 18.7 ± 3.9 \si{\percent} of all sequences (\autoref{gamma}).
Similarly to previous taxa, large taxonomic differences between seawater
and surface associated communities were found. Seawater communities were
mainly comprised of the OM60 (NOR5) clade, \emph{Litoricola},
\emph{Acinetobacter} and the SAR86 clade, while epiphytic communities
were mainly composed of no relative \emph{Gammaproteobacteria} and
\emph{Granulosicocus}. Beside these two groups specific to all three
epiphytic communities, \emph{Cymodocea nodosa} was characterized by
\emph{Arenicella}, no relative \emph{Burkholderiales} and
\emph{Methylotenera}, while \emph{Thioploca}, no relative
\emph{Cellvibrionaceae} and \emph{Reinekea} were more specific to both
invaded and noninvaded \emph{Caulerpa cylindracea}. In addition,
similarly to \emph{Bacteroidota}, a large proportion of the surface
associated community was grouped as other \emph{Gammaproteobacteria}
indicating high diversity within this group (\autoref{gamma}).

@@

@@

@@

@@

@@

@@

@@

>>>>

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of eukaryotic,
chloroplast, mitochondrial and no relative sequences
(\autoref{supp-nseq_notus}). A total of 35 samples originating from
epiphytic archaeal and bacterial communities associated with surfaces of
the seagrass \emph{Cymodocea nodosa} and macroalga \emph{Caulerpa
cylindracea} were analyzed. In addition, 18 samples (one of the samples
was sequenced two times) originating from picoplankton archaeal and
bacterial communities in the surrounding seawater were also processed.
The number of reads per sample ranged between 8,407 and 77,465 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off that is a common
observation in high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences mentioned before reads were clustered into 28,726 different
OTUs at a similarity level of 97 \si{\percent}. Reads numbers were
normalized to the minimum number of sequences, 8,407
(\autoref{supp-nseq_notus}), through rarefaction resulting in 17,007
different OTUs that ranged from 366 to 1,998 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes of richness
and diversity the Observed Number of OTUs, Chao1, ACE, Exponential
Shannon (Jost, 2006) and Inverse Simpson were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{Caulerpa cylindracea} (invaded {[}Number of OTUs,
1,688.5 ± 125.6 OTUs{]} and noninvaded {[}Number of OTUs, 1,744.8 ±
150.8 OTUs{]}), middle vales for \emph{Cymodocea nodosa} (Number of
OTUs, 1,062.8 ± 209.6 OTUs) and lower values for picoplankton
communities in the surrounding seawater (Number of OTUs, 528.1 ± 135.8
OTUs) (\autoref{supp-calculators}). Seasonal changes did not show such
large dissimilarities. Seawater communities richness was stable during
the studied period with the exception of one sampling point in December
2017 when larger values were observed. \emph{Cymodocea nodosa}
communities showed a slow increase towards the end of the study, while
\emph{Caulerpa cylindracea} (invaded and noninvaded) communities were
characterized by slightly larger values in Spring and Summer in
comparison to Autumn and Winter (\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient were, respectively, calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from a
particular environment. The highest proportion of shared OTUs and
community was found between invaded and noninvaded \emph{Caulerpa
cylindracea} (Jaccard, 0.35; Bray-Curtis, 0.77), while lower shared
values were calculated between seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{Cymodocea nodosa}
and \emph{Caulerpa cylindracea} were approximately in the middle between
these two extremes. To assess seasonal changes in the proportion of
shared OTUs and communities the Jaccard's and Bray-Curtis Similarity
Coefficients were calculated between consecutive sampling points
(\autoref{shared}). Both coefficients showed similar trends. Temporal
proportional changes were more pronounced for seawater in comparison to
\emph{Cymodocea nodosa} and especially \emph{Caulerpa cylindracea}
associated communities (\autoref{shared}). To further disentangle the
environmental and seasonal community dissimilarity a Principal
Coordinates Analysis (PCoA) based on Bray-Curtis distances and OTU
abundances was applied. It showed a clear separation between planktonic
and surface associated communities (\autoref{pcoa}). In addition, a
separation of epiphytic bacterial and archaeal communities based on host
species was determined. This separation was further supported by ANOSIM
(R = 0.96, \emph{p} \textless{} 0.001). Seasonal changes of seawater
communities indicated a separation between Spring, Summer and
Autumn/Winter samples (ANOSIM, R = 0.64, \emph{p} \textless{} 0.001).
Epiphytic microbial communities associated with \emph{Cymodocea nodosa}
showed a similar pattern (ANOSIM, R = 0.56, \emph{p} \textless{} 0.01),
while communities from the surfaces of \emph{Caulerpa cylindracea}
indicated a non so strongly supported, as in previous cases, separation
between Summer and Autumn/Winter/Spring samples (ANOSIM, R = 0.31,
\emph{p} \textless{} 0.05) (\autoref{pcoa}).

The taxonomic composition of both, macrophyte associated and seawater
communities, was dominated by \emph{Bacteria} (99.1 ± 2.1 \si{\percent})
over \emph{Archaea} (0.9 ± 2.1 \si{\percent}) (\autoref{community}).
Higher relative abundances of chloroplast related sequences were only
observed in surface associated communities, with higher values in
Autumn/Winter (37.2 ± 11.2 \si{\percent}) in comparison to Spring/Summer
(20.9 ± 9.7 \si{\percent}) (\autoref{supp-chloroplast}). Generally, at
higher taxonomic ranks (phylum-class) epiphytic and seawater microbial
communities were composed of similar bacterial taxa. Seawater
communities were mainly comprised of \emph{Actinobacteriota},
\emph{Bacteroidota}, \emph{Cyanobacteria}, \emph{Alphaproteobacteria},
\emph{Gammaproteobacteria} and \emph{Verrucomicrobiota}. Communities
associated with \emph{Cymodocea nodosa} were consisted of same groups
with the addition of \emph{Planctomycetota} whose contribution was
higher in summer 2018. In addition, communities from invaded and
noninvaded \emph{Caulerpa cylindracea} were similar and characterized by
same groups as seawater and \emph{Cymodocea nodosa} communities with the
addition of \emph{Desulfobacterota} (\autoref{community}). Larger
differences between environments and host species could be observed at
lower taxonomic ranks (\autoref{cyano} -- \ref{desulfo}).

\emph{Cyanobacteria} related sequences were comprising, on average, 5.5
± 4.4 \si{\percent} of total sequences (\autoref{cyano}). Higher
proportions were found for \emph{Cymodocea nodosa} (16.4 ± 5.3
\si{\percent}) and \emph{Caulerpa cylindracea} (invaded {[} (7.7 ± 3.9
\si{\percent}){]} and noninvaded {[} (7.8 ± 2.4 \si{\percent}){]})
associated communities in autumn and for seawater communities in winter
(8.8 ± 7.4 \si{\percent}). Large taxonomic differences between surface
associated and seawater cyanobacterial communities were observed.
Seawater communities were mainly comprised of \emph{Cyanobium} and
\emph{Synechococcus}, while surface associated communities were
consisted of \emph{Pleurocapsa} and sequences without known relatives
within \emph{Cyanobacteriia} (\autoref{cyano}). In addition, seasonal
changes in surface associated communities were observed with
\emph{Pleurocapsa} and no relative \emph{Cyanobacteriia} comprising
larger proportions in autumn and winter and \emph{Acrophormium},
\emph{Phormidesmis} and no relative \emph{Nodosilineaceae} in spring and
summer (\autoref{cyano}).

Sequences classified as \emph{Bacteroidota} were comprising, on average,
19.2 ± 5.5 \si{\percent} of all sequences (\autoref{bactero}). Similarly
to \emph{Cyanobacteria}, large differences in the taxonomic composition
between seawater and surface associated communities were found
(\autoref{bactero}). The seawater community was characterized by the NS4
and NS5 marine groups, uncultured \emph{Cryomorphaceae}, uncultured
\emph{Flavobacteriaceae}, NS11-12 marine group, \emph{Balneola},
uncultured \emph{Balneolaceae} and \emph{Formosa}. In contrast, in
surface associated communities \emph{Lewinella}, \emph{Portibacter},
\emph{Rubidimonas}, no relative \emph{Saprospiraceae}, uncultured
\emph{Saprospiraceae}, no relative \emph{Flavobacteriaceae} and
uncultured \emph{Rhodothermaceae} were found. Some groups showed slight
seasonal changes such as no relative \emph{Flavobacteriaceae} that were
more pronounced from November 2017 until June 2018. In contrast,
uncultured \emph{Rhodothermaceae} showed higher proportions from June
2018 until the end of the study period. Surface associated
\emph{Bacteroidota} communities were very diverse as could be observed
in the the high proportion of taxa that grouped as other
\emph{Bacteroidota} (\autoref{bactero}).

On average, \emph{Alphaproteobacteria} were in comparison to other high
rank taxa the largest taxonomic group, comprising 29.2 ± 12.0
\si{\percent} of all sequences (\autoref{alpha}). In accordance to
previous taxa, high differences between seawater and surface associated
communities were observed. Picoplankton communities were composed mainly
of the SAR11 clade, AEGEAN-169 marine group, SAR116 clade, no relative
\emph{Rhodobacteraceae}, HIMB11 and OCS116 clade, while surface
associated communities were composed in high proportion of no relative
\emph{Rhodobacteraceae} and to a lesser degree of \emph{Pseudoahrensia},
no relative \emph{Alphaproteobacteria}, no relative
\emph{Hyphomonadaceae} and \emph{Amylibacter}. Representatives of no
relative \emph{Rhodobacteraceae} were comprising on average 40.6 ± 23.2
\si{\percent} of all alphaproteobacterial sequences from the epiphytic
community (\autoref{alpha}). In addition, \emph{Amylibacter} was
detected mainly in \emph{Cymodocea nodosa} from November 2017 until
March 2018.

Sequences related to \emph{Gammaproteobacteria} were comprising, on
average, 18.7 ± 3.9 \si{\percent} of all sequences (\autoref{gamma}).
Similarly to previous taxa, large taxonomic differences between seawater
and surface associated communities were found. Seawater communities were
mainly comprised of the OM60 (NOR5) clade, \emph{Litoricola},
\emph{Acinetobacter} and the SAR86 clade, while epiphytic communities
were mainly composed of no relative \emph{Gammaproteobacteria} and
\emph{Granulosicocus}. Beside these two groups specific to all three
epiphytic communities, \emph{Cymodocea nodosa} was characterized by
\emph{Arenicella}, no relative \emph{Burkholderiales} and
\emph{Methylotenera}, while \emph{Thioploca}, no relative
\emph{Cellvibrionaceae} and \emph{Reinekea} were more specific to both
invaded and noninvaded \emph{Caulerpa cylindracea}. In addition,
\emph{Arenicella} was more pronounced in November and December 2017,
while no relative \emph{Burkholderiales} and \emph{Methylotenera} were
more characteristic for the period form March until May 2018. For the
\emph{Caulerpa cylindracea} specific taxa no relative
\emph{Cellvibrionaceae} and \emph{Reinekea} showed some seasonality and
were characterisitic for samples originating from June to October 2018.
In addition, similarly to \emph{Bacteroidota}, a large proportion of the
surface associated community was grouped as other
\emph{Gammaproteobacteria} indicating high diversity within this group
(\autoref{gamma}).

@@

Surfaces of marine macrophytes are harboring biofilms consisted of
diverse microbial taxa (Egan \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019). No standard protocol has been developed to study these
macophyte associated microbes (Ugarelli \emph{et al.}, 2019). Different
procedures of microbial cells removal from host surfaces were described,
such as host tissue shaking (Nõges \emph{et al.}, 2010), scraping (Uku
\emph{et al.}, 2007) and ultrasonication (Cai \emph{et al.}, 2014). All
these methods showed different removal efficiencies and none was
enabling a complete removal of attached microbial cells. In the present
study, we applied an earlier developed removal protocol (Korlević
\emph{et al.}, submitted), based on a previous idea of direct cellular
lysis (Burke \emph{et al.}, 2009), to ensure an almost complete cell
detachment. The application of a direct lysis procedure coupled with a
high frequency sampling protocol and Illumina high resolution amplicon
sequencing has enabled us to make a detailed description of bacterial
and archaeal communities associated with the surfaces of two marine
macrophytes, \emph{Cymodocea nodosa} and \emph{Caulerpa cylindracea}.

Observed highest richness values for \emph{Caulerpa cylindracea}
(invaded and noninvaded), middle for \emph{Cymodocea nodosa} and lowest
for seawater derived communities. Higher values for surface associated
communities in comparison to seawater were described earlier for
seagrasses (Ugarelli \emph{et al.}, 2019) and could be attributed to a
larger set of inhabitable microniches existing on macrophyte surfaces.
In addition, highest values observed for \emph{Caulerpa cylindracea} are
probably a consequence of sediment derived OTUs that were present only
in \emph{Caulerpa cylindracea} communities. \emph{Caulerpa cylindracea}
stolon is attached to surface sediment with rhizoids, so the stolon and
rhizoids are in a direct contact with the sediment surface. Part of the
surface attached \emph{Caulerpa cylnidracea} community is therefore
comprised of sediment derived cells that could cause the observed
increase in richness. In addition, seasonal richness differences
observed for surface attached communities showed slightly higher values
in spring and summer could be explained by a higher macrophyte growth in
these seasons (M. Najdek, personal communication; Zavodnik \emph{et
al.}, 1998; Ruitton \emph{et al.}, 2005). During active periods
macrophytes exhibit a more dynamic chemical interaction with the surface
community causing an opening of a larger set of microniches.

@@

\leavevmode\hypertarget{ref-Burke2009}{}%
Burke, C., Kjelleberg, S., and Thomas, T. (2009) Selective extraction of
bacterial DNA from the surfaces of macroalgae. \emph{Applied and
environmental microbiology} \textbf{75}: 252--256.

@@

\leavevmode\hypertarget{ref-Cai2014}{}%
Cai, X., Gao, G., Yang, J., Tang, X., Dai, J., Chen, D., and Song, Y.
(2014) An ultrasonic method for separation of epiphytic microbes from
freshwater submerged macrophytes. \emph{Journal of basic microbiology}
\textbf{54}: 758--761.

@@

\leavevmode\hypertarget{ref-Noges2010}{}%
Nõges, T., Luup, H., and Feldmann, T. (2010) Primary production of
aquatic macrophytes and their epiphytes in two shallow lakes (Peipsi and
Võrtsjärv) in Estonia. \emph{Aquatic Ecology} \textbf{44}: 83--92.

@@

\leavevmode\hypertarget{ref-Ruitton2005}{}%
Ruitton, S., Verlaque, M., and Boudouresque, C.F. (2005) Seasonal
changes of the introduced Caulerpa racemosa var. cylindracea
(Caulerpales, Chlorophyta) at the northwest limit of its Mediterranean
range. \emph{Aquatic Botany} \textbf{82}: 55--70.

@@

\leavevmode\hypertarget{ref-Uku2007}{}%
Uku, J., Björk, M., Bergman, B., and Díez, B. (2007) Characterization
and comparison of prokaryotic epiphytes associated with three East
African seagrasses. \emph{Journal of Phycology} \textbf{43}: 768--779.

@@

\leavevmode\hypertarget{ref-Zavodnik1998}{}%
Zavodnik, N., Travizi, A., and De Rosa, S. (1998) Seasonal variations in
the rate of pliotosynthetic activity and chemical composition of the
seagrass Cymodocea nodosa (Ucr.) Asch. \emph{Scientia Marina}
\textbf{62}: 301--309.
